---
title: "Baltimore County 911 Overdose Calls Analysis"
author: "Raphael Romero Ruiz"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Z-Score Analysis of Baltimore County 911 Calls

## Introduction

In this analysis, we'll examine data on 911 drug overdose calls from Baltimore County to identify unusually busy or quiet days using z-scores. By calculating the number of calls per day and analyzing their statistical distribution, we can determine which days experienced an abnormal volume of emergency calls - information that could lead to potential story ideas.

A z-score measures how many standard deviations a data point is from the mean. Generally:
- Z-scores between -2 and 2 represent typical values (about 95% of the data)
- Z-scores beyond ±2 indicate unusual values that warrant investigation
- Z-scores beyond ±3 are very unusual (only about 0.3% of normally distributed data)

## Task 1: Setup and Data Loading

Let's start by loading the necessary packages and our dataset:

```{r load-packages}
library(tidyverse)  # For data manipulation and visualization
library(lubridate)  # For handling dates
library(knitr)      # For nice tables
library(scales)     # For better axis formatting
```

```{r load-data}
# Load the 911 calls data
calls_data <- read_csv("https://raw.githubusercontent.com/dwillis/datajournalismbook-maryland/refs/heads/main/data/baltco_911_calls.csv")

# Display the first few rows to understand the structure
head(calls_data)
```

## Task 2: Initial Data Exploration

Let's explore the basic structure and characteristics of our dataset:

```{r explore-data}
# Get a summary of the data
glimpse(calls_data)

# Check for any missing values
colSums(is.na(calls_data))

# Check the date range in our data
min_date <- min(calls_data$date)
max_date <- max(calls_data$date)
cat("Data ranges from", min_date, "to", max_date)

```

## Task 3: Aggregating Calls by Day (3 points)

Now let's count how many 911 overdose calls occurred on each day:

```{r aggregate-by-day}
# Count calls per day
daily_calls <- calls_data |>
  count(date, name = "call_count")

# Display the first few days
head(daily_calls, 10) |>
  kable()

# Basic summary statistics for daily call volume
summary(daily_calls$call_count)
```

Let's visualize the time series of daily calls. What patterns do you see? If you had to draw horizontal lines that would highlight unusual days, where would those lines be?

There's a high amount of calls in the late spring 2022 where there about on average 10 calls. There are some obvious outliers like the high number of calls on a date in October 2022 - the peak in our data. 

```{r daily-calls-plot}
# Plot daily call volume
ggplot(daily_calls, aes(x = date, y = call_count)) +
  geom_line(color = "darkblue", alpha = 0.7) +
  geom_point(color = "darkblue", alpha = 0.5, size = 1) +
  theme_minimal() +
  labs(title = "Daily 911 Overdose Calls in Baltimore County",
       x = "Date",
       y = "Number of Calls",
       caption = "Source: Baltimore County 911 Call Data") +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




## Task 4: Calculating Z-Scores for Daily Call Volume

Now, let's calculate z-scores to identify statistically unusual days:

```{r calculate-z-scores}
# Calculate z-scores for daily call counts
daily_calls <- daily_calls |>
  mutate(
    mean_calls = mean(call_count),
    sd_calls = sd(call_count),
    z_score = (call_count - mean_calls) / sd_calls
  )

# Display the data with z-scores
daily_calls |>
  select(date, call_count, z_score) |>
  head(10) |>
  kable(digits = 2)
```

## Task 5: Identifying Unusual Days (3 points)

Let's identify days with unusually high or low call volumes (|z-score| > 2):

```{r identify-unusual-days}
# Find days with unusual call volumes
unusual_days <- daily_calls |>
  filter(abs(z_score) > 2) |>
  arrange(desc(abs(z_score)))

# Display unusual days
unusual_days |>
  select(date, call_count, z_score) |>
  kable(digits = 2)

# Count how many days were unusually busy vs. unusually quiet
unusual_days |>
  group_by(unusual_type = if_else(z_score > 0, "Unusually High", "Unusually Low")) |>
  summarise(count = n()) |>
  kable()
```

Let's visualize these unusual days in the context of the full dataset. How do they differ from your answer on Task 3?

ANSWER HERE

```{r plot-unusual-days}
# Create a clean factor for the plot legend
daily_calls <- daily_calls |>
  mutate(call_volume_type = case_when(
    z_score > 2 ~ "Unusually High",
    z_score < -2 ~ "Unusually Low",
    TRUE ~ "Normal"
  ))

# Plot with unusual days highlighted
ggplot(daily_calls, aes(x = date, y = call_count, color = call_volume_type)) +
  geom_line(color = "gray70") +
  geom_point(size = 2) +
  scale_color_manual(values = c("Normal" = "gray50", 
                               "Unusually High" = "red", 
                               "Unusually Low" = "blue")) +
  theme_minimal() +
  labs(title = "Daily 911 Overdose Calls with Unusual Days Highlighted",
       subtitle = "Unusual days have z-scores with absolute values > 2",
       x = "Date",
       y = "Number of Calls",
       color = "Call Volume",
       caption = "Source: Baltimore County 911 Call Data") +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Task 6: Day of Week Patterns

Let's also check if certain days of the week typically see more 911 calls:

```{r day-of-week-analysis}
# Add day of week to our dataset
daily_calls <- daily_calls |>
  mutate(
    day_of_week = wday(date, label = TRUE)
  )

# Calculate average calls by day of week
dow_summary <- daily_calls |>
  group_by(day_of_week) |>
  summarise(
    avg_calls = mean(call_count),
    median_calls = median(call_count),
    min_calls = min(call_count),
    max_calls = max(call_count)
  ) |>
  arrange(match(day_of_week, c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Display summary by day of week
dow_summary |> kable(digits = 1)

# Plot calls by day of week
ggplot(dow_summary, aes(x = day_of_week, y = avg_calls)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(avg_calls, 1)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Average 911 Overdose Calls by Day of Week",
       x = "Day of Week",
       y = "Average Number of Calls",
       caption = "Source: Baltimore County 911 Call Data") +
  theme(axis.text.x = element_text(angle = 0))
```

## Task 7: Accounting for Day of Week Pattern (3 points)

We can also calculate z-scores that account for the day of the week pattern. How would you describe the day with the highest z-score without using the term "z-score"?

ANSWER HERE

```{r day-of-week-z-scores}
# Calculate day-of-week adjusted z-scores
daily_calls <- daily_calls |>
  group_by(day_of_week) |>
  mutate(
    dow_mean = mean(call_count),
    dow_sd = sd(call_count),
    dow_z_score = (call_count - dow_mean) / dow_sd
  ) |>
  ungroup()

# Find days that were unusual for their day of the week
dow_unusual_days <- daily_calls |>
  filter(abs(dow_z_score) > 2) |>
  arrange(desc(abs(dow_z_score)))

# Display days that were unusual for their day of the week
dow_unusual_days |>
  select(date, day_of_week, call_count, dow_mean, dow_z_score) |>
  kable(digits = 2)
```


## Task 8: Temporal Patterns and Trends

Let's also look for seasonal patterns or trends over time:

```{r temporal-analysis}
# Add month and year for seasonal analysis
daily_calls <- daily_calls |>
  mutate(
    month = month(date, label = TRUE),
    year = year(date)
  )

# Calculate average calls by month
monthly_summary <- daily_calls |>
  group_by(month) |>
  summarise(
    avg_calls = mean(call_count),
    median_calls = median(call_count)
  )

# Plot calls by month
ggplot(monthly_summary, aes(x = month, y = avg_calls, group = 1)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Average 911 Overdose Calls by Month",
       x = "Month",
       y = "Average Number of Calls",
       caption = "Source: Baltimore County 911 Call Data")

# Show a rolling 7-day average to better see trends
daily_calls <- daily_calls |>
  arrange(date) |>
  mutate(rolling_avg_7day = zoo::rollmean(call_count, k = 7, fill = NA))

# Plot with 7-day rolling average
ggplot(daily_calls, aes(x = date)) +
  geom_line(aes(y = call_count), color = "gray60", alpha = 0.7) +
  geom_line(aes(y = rolling_avg_7day), color = "blue", size = 1) +
  theme_minimal() +
  labs(title = "Daily 911 Overdose Calls with 7-Day Moving Average",
       x = "Date",
       y = "Number of Calls",
       caption = "Source: Baltimore County 911 Call Data") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Task 9: T Most Unusual Days

For the most unusual days, we might want to check if they coincide with significant events:

```{r investigate-top-outliers}
# Get the top 10 most unusual days
top_outliers <- daily_calls |>
  arrange(desc(abs(z_score))) |>
  head(10) |>
  select(date, day_of_week, call_count, z_score)

# Display these days
top_outliers |> kable(digits = 2)
```

## Task 10: Journalistic Insights (6 points)

Based on our z-score analysis, what are some potential angles for reporting? What would your next steps be and what would you tell an editor about your findings?
```{r cbp-drug-seizure-data}
cbp_data_22    <- read_csv("https://www.cbp.gov/sites/default/files/2025-11/nationwide-drugs-fy22-fy25.csv")
cbp_data_21 <- read_csv("https://www.cbp.gov/sites/default/files/2024-10/nationwide-drugs-fy21-fy24.csv")
```

```{r cbp-combine}
# stack both files and drop any rows that appear in both
cbp_combined <- bind_rows(cbp_data_22, cbp_data_21) |>
  distinct()
```

```{r cbp-monthly}
# pull only fentanyl and heroin from the southwest border,
# then convert the fiscal year + month to a real calendar date
# (CBP's fiscal year starts in October, so Oct-Dec belong to the prior calendar year)
cbp_monthly <- cbp_combined |>
  filter(`Drug Type` %in% c("Fentanyl", "Heroin")) |>
  filter(Region == "Southwest Border") |>
  mutate(
    fy_num     = as.integer(FY),
    month_num  = match(`Month (abbv)`,
                       c("JAN","FEB","MAR","APR","MAY","JUN",
                         "JUL","AUG","SEP","OCT","NOV","DEC")),
    cal_year   = if_else(month_num >= 10L, fy_num - 1L, fy_num),
    year_month = make_date(cal_year, month_num, 1)
  ) |>
  group_by(year_month) |>
  summarise(total_seizures = sum(`Count of Event`, na.rm = TRUE)) |>
  filter(year_month >= min_date & year_month <= max_date) |>
  mutate(z_seizures = (total_seizures - mean(total_seizures)) / sd(total_seizures))
```

```{r calls-monthly}
# roll up daily calls to monthly averages to match the CBP data's granularity
monthly_calls <- daily_calls |>
  mutate(year_month = make_date(year, as.integer(month), 1)) |>
  group_by(year_month) |>
  summarise(avg_calls = mean(call_count)) |>
  filter(year_month >= min_date & year_month <= max_date) |>
  mutate(z_calls = (avg_calls - mean(avg_calls)) / sd(avg_calls))
```

```{r overdose-seizure-plot}
ggplot() +
  geom_line(data = monthly_calls,
            aes(x = year_month, y = z_calls, color = "Balt. County 911 Overdose Calls")) +
  geom_point(data = monthly_calls,
             aes(x = year_month, y = z_calls, color = "Balt. County 911 Overdose Calls"), size = 2) +
  geom_line(data = cbp_monthly,
            aes(x = year_month, y = z_seizures, color = "CBP Fentanyl/Heroin Seizures"),
            linetype = "dashed") +
  geom_point(data = cbp_monthly,
             aes(x = year_month, y = z_seizures, color = "CBP Fentanyl/Heroin Seizures"), size = 2) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray40") +
  scale_color_manual(values = c(
    "Balt. County 911 Overdose Calls" = "darkblue",
    "CBP Fentanyl/Heroin Seizures"    = "red"
  )) +
  theme_minimal() +
  labs(
    title    = "Balt. County Overdose Calls vs. Southwest Border Seizures",
    subtitle = "Both series normalized to z-scores — dotted line = average",
    x        = "Date",
    y        = "Z-Score",
    color    = "Data Source",
    caption  = "Sources: Baltimore County 911 Data; CBP Southwest Border"
  ) +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The z-score analysis points to some clear story angles that I'd pursue. The most striking to me is that Baltimore County overdose calls were unusually high throughout the spring and summer of 2022, with August hitting a peak of 366 calls. And when I compared this to CBP Southwest Border seizure data, both lines peaked at roughly the same time.

My question is, if border seizures were truly choking off the supply reaching the streets, we’d see an inverse relationship: more drugs caught, fewer overdoses. But the data shows the opposite. Why? Both of the lines (which I plotted above) trend together. This may suggest that seizure counts measure total trafficking volume rather than interdiction success. When more fentanyl moves through the border, but at the same time more is caught and more is getting through.

The September 2022 divergence, where 911 calls dropped 21% and border drug seizures stayed elevated—adds another layer of complexity. Local overdose volume clearly isn't just dictated by illicit border activity. If I were to report this out, my next step would be to pull 2024 and 2025 911 data to see if this trend holds. I also would want to ask CBP directly what percentage of estimated trafficking volume their seizure numbers represent, especially since March 2022 hit a record 171 seizure events just as Baltimore County calls were spiking. Most importantly, I need to talk to Baltimore County narcotics detectives. They’ll have the ground-level read on whether these border surges actually translate to changes in local supply or if the street-level picture is driven by something else entirely.