
---
title: "Multiple Regression Analysis in R"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships, such as how different factors collectively impact outcomes like election results, economic indicators, or sports performances. By understanding these relationships, journalists can provide deeper insights and more accurate predictions.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Data Import

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains various statistics for NFL teams from the 2019 season. This dataset will be used to build our model.

```{r}
teams <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/main/nfl_2019.csv")
```

## Model Building

Next, we construct a linear model to analyze how different factors like `Yards Gained`, `Takeaways`, and `Points Scored` influence the number of wins. This model helps us quantify the relationships and assess which factors are significant predictors of success.

```{r}
model <- lm(Wins ~ `Yards Gained` + Takeaways + `Points Scored`, data = teams)
summary(model)
```

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For your model:

*   **Min and Max** values indicate the range of residuals. A maximum of 4.1361 and a minimum of -4.3780 suggest that there might be some substantial errors in predictions for some cases, but generally, the errors are relatively evenly distributed around zero.
*   **Median** close to zero (-0.0604) indicates that the model is generally centered around the actual data points, but still, some deviations exist.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

*   **Yards Gained**: The coefficient is -0.001329. This suggests a very small negative impact of yards gained on wins, but this effect is not statistically significant (p = 0.34332), implying that as yards gained increase, the number of wins slightly decreases, which is counterintuitive and likely indicates no real effect or that the model isn't capturing the complexities.
*   **Takeaways**: The coefficient is 0.150593, suggesting that each additional takeaway is associated with an increase in wins by about 0.151, which is near the threshold of significance (p = 0.07928). This suggests a potentially meaningful, though not conventionally statistically significant, positive effect on wins.
*   **Points Scored**: The coefficient is 0.038732, which is significant (p = 0.00665). This indicates that each additional point scored is associated with an increase in wins by about 0.039, suggesting a clear and meaningful impact.

### Model Fit (3 points)

*   **Residual Standard Error**: The RSE of 2.072 tells us that the typical prediction by the model deviates from the actual number of wins by about 2.072 wins.
*   **Multiple R-squared**: 0.6232 indicates that about 62.32% of the variability in wins is explained by the model, which is relatively high, suggesting a good fit.
*   **Adjusted R-squared**: 0.5828 adjusts the R-squared for the number of predictors and indicates that about 58.28% of the variance is explained by the model after accounting for the number of variables used.
*   **F-statistic**: The F-statistic of 15.44 with a very low p-value (4.058e-06) suggests the model as a whole is statistically significant.

Is this a useful model, based on what you've seen? Why?

ANSWER: Yes, but with caveats. An R-squared of 0.6232 means it accounts for about 62% of the variation in wins across teams, even though NFL outcomes are also shaped by coaching decisions, injuries, weather, and special teams play that these three variables cannot capture. Points Scored is the only variable that clears the significance threshold (p = 0.007), meaning it is the only one we can confidently say actually predicts wins. Where it falls short is precision. The residual standard error of roughly 2 wins means predictions can be off by two games in either direction in a 17-game season. That is too wide a margin to publish a win-total prediction and stand behind it. The model is more useful as a reporting tool than a forecasting tool. It tells you which statistics actually correlate with winning so you know what to focus on and what to ignore when covering a team.

## Correlation Analysis

After building the model, we calculate the correlation matrix for our variables of interest. This matrix helps visualize the strength and direction of the relationships between each pair of variables.

```{r}
selected_vars <- teams |>  select(`Yards Gained`, Takeaways, `Points Scored`, Wins)
cor(selected_vars)
```

## Visualization of Correlation Matrix

Finally, we use a visualization tool to plot the correlation matrix. This graphical representation makes it easier to interpret the correlations at a glance, showing which variables are most closely connected.

```{r}
corr_matrix <- cor(selected_vars)
corrplot(corr_matrix, method = "circle")
```

## Results Interpretation (4 points)

The results from our correlation analysis show the strength of the relationships between the number of wins and factors like yards gained, takeaways, and points scored. High positive correlations indicate strong relationships where increases in one variable are associated with increases in another. Conversely, negative correlations suggest an inverse relationship. By understanding these dynamics, journalists can better explain the factors that contribute to a team's success or failure in a season.

Which of these three factors - Yards Gained, Takeaways and Points Scored - have the most impact on wins? Are there any that do not seem impactful at all?

ANSWER: Of the three, Points Scored drives wins most reliably. It is the only variable with a statistically significant coefficient (p = 0.007), and the correlation matrix confirms it has the strongest positive relationship with wins of the three. For a reporter, this means scoring output is the most defensible stat to anchor a team evaluation on. Takeaways are worth tracking as a secondary signal. The relationship approaches significance (p = 0.079) and the direction is positive, so turnover-forcing defenses likely do win more, but the data does not quite get there. Yards Gained is the most surprising finding and arguably the most useful one journalistically. Its coefficient is near zero and its p-value (0.343) clears no meaningful bar, which means a team racking up big yardage numbers but losing games is not being betrayed by bad luck. The yards just do not matter as much as the scoreboard does. That is a story worth telling.

## Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this model to cover a team or game.

ANSWER: You could use this model to find teams whose records do not match what their stats predict, and then report on why. A team that scores a lot of points and forces turnovers but sits at .500 is underperforming a model-based expectation. That gap is a story. Maybe it is special teams. Maybe it is clock management. On the flip side, a team with a winning record but poor scoring numbers is probably winning close games on borrowed time, and that is worth writing about too. Midseason is the most natural time to do this. Once a team has eight or ten games of data, you can run their numbers through the model and ask whether their record reflects how they are actually playing. If the numbers say 9-4 and they are 6-7, you have a concrete statistical frame for a piece about a team that is worse than their potential. It grounds the analysis in something more rigorous than a beat reporter's eye test.
