---
title: "Hypothesis Testing & Effect Size"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Did Community Policing Actually Improve Response Times?

The city of Westbrook implemented a new community policing program in January. The police chief claims the program has improved emergency response times. Before the program, the average response time was **8.2 minutes** with a standard deviation of **1.5 minutes**. You've collected data from 20 randomly selected emergency calls after the program started.

Your editor wants to know two things: Is the improvement real? And is it big enough to matter?

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Set up the data and the hypotheses (2 points)

Before we run any tests, we need to state our hypotheses. The null hypothesis is what we assume to be true until we have evidence otherwise — like "innocent until proven guilty." Fill in the REPLACE_ME values.

**Null Hypothesis (H₀):** The response times have not changed.

**Alternative Hypothesis (H₁):** The response times have changed, has dropped the response times.

```{r}
# Response times (in minutes) after the community policing program
response_times <- c(
  7.8, 8.1, 7.5, 7.9, 8.3,
  7.6, 7.4, 8.0, 7.7, 7.8,
  8.2, 7.3, 7.9, 7.5, 8.1,
  7.6, 7.8, 7.4, 7.7, 7.9
)

# Known baseline (before the program)
baseline_mean <- 8.2
baseline_sd <- 1.5

# Create a tibble
response_data <- tibble(
  call_number = 1:20,
  response_time = response_times
)

response_data
```

### Task 3: Calculate summary statistics (1 point)

Let's see what the new response times look like. Replace the REPLACE_ME values.

```{r}
summary_stats <- response_data |>
  summarize(
    sample_size = n(),
    sample_mean = mean(response_time),
    sample_sd = sd(response_time),
    standard_error = sample_sd / sqrt(sample_size)
  )

summary_stats
```

**Answer this question:** The sample mean is lower than the baseline of 8.2 minutes. How much lower? Does that seem like a meaningful improvement to you, just based on the numbers? What would you want to know before deciding?

The new sample_mean is telling us that response times on average are now at about 7.8 minutes, which is about 0.4 minute drop. This seems statistcally too similar for the change to be important for us to report on. However we should look at how much the standard deviation has changed now sitting at 0.28. This means our data is much tighter so the range of time from these response calls don't vary as much anymore. 

### Task 4: Perform a one-sample t-test (2 points)

Now we formally test the police chief's claim. Since we're testing whether response times *decreased*, we use a one-sided test (`alternative = "less"`). Replace the REPLACE_ME with the baseline mean we're comparing against.

```{r}
# Set our significance level
alpha <- 0.05

# Perform the t-test
t_test_result <- t.test(
  response_data$response_time,
  mu = 8.2,
  alternative = "less"
)

t_test_result
```

### Task 5: Interpret the t-test results (2 points)

Let's pull out the key numbers. Look at the p-value: is it less than our alpha of 0.05?

```{r}
# Extract key values
p_value <- t_test_result$p.value

results <- tibble(
  `Baseline mean` = paste(baseline_mean, "minutes"),
  `New mean` = paste(round(summary_stats$sample_mean, 2), "minutes"),
  `Difference` = paste(round(baseline_mean - summary_stats$sample_mean, 2), "minutes"),
  `p-value` = round(p_value, 6),
  `Significant?` = if_else(p_value < alpha, "Yes - reject null hypothesis", "No - fail to reject")
)

results
```

**Answer this question:** Based on the p-value, can we reject the null hypothesis? What does this tell the police chief — and what does it NOT tell him?

Well we can certainly say that some change in the response times has happened. We can not immediately agree with the police chief to say that the response times have improved quiet yet. 

### Task 6: Visualize the results

```{r}
ggplot(response_data, aes(x = response_time)) +
  geom_histogram(binwidth = 0.2, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = baseline_mean, color = "red", linetype = "dashed", size = 1.2) +
  geom_vline(xintercept = summary_stats$sample_mean, color = "darkgreen", size = 1.2) +
  annotate("text", x = baseline_mean + 0.05, y = 4,
           label = paste("Baseline:", baseline_mean),
           color = "red", hjust = 0, fontface = "bold") +
  annotate("text", x = summary_stats$sample_mean - 0.05, y = 3.5,
           label = paste("New Mean:", round(summary_stats$sample_mean, 2)),
           color = "darkgreen", hjust = 1, fontface = "bold") +
  labs(
    title = "Emergency Response Times: Before vs. After Community Policing",
    subtitle = paste0("p-value = ", round(p_value, 6)),
    x = "Response Time (minutes)",
    y = "Number of Calls"
  ) +
  theme_minimal()
```

### Task 7: Calculate the effect size (3 points)

The t-test told us the improvement is statistically significant — it's real, not due to chance. But is it *meaningful*? That's what effect size measures. We'll calculate **Cohen's d**, which tells us how large the difference is in standard deviation units.

The formula is: d = (baseline mean - sample mean) / baseline standard deviation

Replace the REPLACE_ME values.

```{r}
effect_size <- (baseline_mean - summary_stats$sample_mean) / baseline_sd

effect_size_interpretation <- case_when(
  abs(effect_size) < 0.2 ~ "Negligible",
  abs(effect_size) < 0.5 ~ "Small",
  abs(effect_size) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `Cohen's d` = round(effect_size, 3),
  `Interpretation` = effect_size_interpretation,
  `Difference in minutes` = round(baseline_mean - summary_stats$sample_mean, 2),
  `Difference in seconds` = round((baseline_mean - summary_stats$sample_mean) * 60, 0)
)
```

**Answer this question:** The result is statistically significant, but the effect size is small. In practical terms, the improvement is about 25 seconds per call. The police chief is claiming success. Would you write a story saying the program works? How would you frame the findings honestly — acknowledging both that the improvement is real AND that it's small?

This to me does not scream success. This feels more like Bush hopping on a military carrier with a banner reading "Mission Accomplished" in the back. There is no denying that a change, one that drops the response time calls, is happening. 

But putting the statistical findings in seconds really helps us understand just how small that "improvement" actually is. If I was writing a story about this, I would question the Police Chief's zealous claim and ask him how this has a real world impact on fighting crime if his officer's are only a few seconds earlier on the scene than before. I think getting a better sense of how the improvement came to be might help shine some light on issues that are being solved or tackled by the department to tackle bad response call numbers.

But we still have to acknowledge that more than 5 minutes still go by after a call is made by the time a crime is reported. In order to frame this better, I'd like to get some criminal justice academic's voices involved on how much of an improvement shaving a few seconds off the clock is in this context. Maybe if we had the time to find some data about the duration a crime event lasts would also help us frame the response time better as well. I say that to mean like, if this is a hostage bank heist sitaution a few seconds may matter in how police are able to catch perpetrators. But, say in a mugging situation, I'm not sure showing up on scene just a few seconds short than than the previous average response time of over 8 minutes is much help at all. 

### Task 8: The journalist's framework (5 points)

The videos you watched describe two questions you need to ask about any statistical result:

1. **Is it statistically significant?** (Is the result unlikely to be due to chance?)
2. **Is it meaningfully significant?** (Is the effect large enough to matter?)

Apply this framework to the community policing data:

**Question 1:** Write 1-2 sentences answering whether the result is statistically significant and how you know.

Our p-value in this case is really small, which means our null hypothesis has to be rejected. It also tells us that the change is unlikely to be due to a random error. Off that we can assume some significance to the data. Furthermore, this is statistically significant change evident by the drop in the standard deviation which means that time range a response call takes has squeezed making the data less noisy. The average response call time, however small, has changed. 

**Question 2:** Write 1-2 sentences answering whether the result is meaningfully significant. Consider: the program cost the city $2 million to implement. Is a 25-second improvement in response times worth that investment? What context would you need to answer that question?

For a cost of $2M, we have to really scrutinize whether the public funds were put to good use here. We need more information to do that including the raw number of calls for service, or at the very least a larger sample size, to see if we are tackling this issue with an appropriate amount of capital. We can use that to compare calls for service by the town's population (in a per capita approach) and compare how other cities with fare. With a per capita approach we can look across the country to see how our own department is performing and whether the city is paying enough money to solve this issue. 

**Question 3:** Write a headline and a 2-3 sentence lede for a story about these findings. Your text should accurately reflect both the statistical significance and the effect size — don't let the police chief's framing drive the story if the data tells a more nuanced one.

**Headline:** Westbrook's $2M Policing Overhaul Trimmed Response Times by 25 Seconds

**Lede:** Westbrook's police chief declared his department's $2 million community policing overhaul a success — but data from 20 emergency calls tells a more complicated story. The improvement amounts to roughly 25 seconds per call. The department has not said how many total calls the program affected, or how Westbrook's numbers compare to similar cities.


When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
