```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

## Does Poverty Predict Math Scores? Testing a School Board Member's Claim

### Introduction

A school board member in Montgomery County, Pennsylvania has claimed at a public meeting that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests. That's a testable claim, and as data journalists, our job is to test it.

We have school-level data from Montgomery County, Pa. that includes each school's FARMS percentage (the share of students eligible for Free and Reduced-price Meal System) and the percentage of students who met or exceeded math expectations on state tests. We'll use linear regression to evaluate whether the school board member's claim holds up, and we'll learn something important about what happens when you trust a model before examining your data.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_pa.csv")
```

### Task 1: Examine the Data Before You Plot Anything

Before we visualize anything, let's look at what we're working with.

```{r}
summary(schools$`FARMS Pct`)

schools |> arrange(desc(`FARMS Pct`)) |> head(10)

schools |> arrange(`FARMS Pct`) |> head(10)
```

Look at the distribution of FARMS percentages. Do any values look unusual or suspicious? What might it mean if a school has a FARMS percentage of exactly 100? Write two or three sentences describing what you see and whether anything concerns you.

There are eight schools with a FARMS percentage of exactly 100. This is a stat I'd want to confirm with the district about just because having literally every student eligible for free or reduced lunch seems unusual and could be something to dig into with more reporting. These schools would represent the population of student in Montgommery County with the most vulnerable kids in the community. 

The biggest finding from this transformation is that we can prove the Montgomery County school board member's point about schools with a majority of students on FARMS have lower test rates. These students test between 57 and 78 percent meeting expectations.


### Task 2: Create a Scatterplot and Describe the Results

Now let's visualize the relationship between FARMS percentage and math scores for all schools.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County PA")
```

Describe what you see. Is there a clear trend? Does the regression line look like it represents the data well, or does it seem to be pulled in an odd direction? Connect what you observed in Task 1 to what you see here.

There's no clear trend here most of the points scatter widely across all FARMS values with no obvious pattern. The regression line is pretty flat with a slight positive slope, showing us how those eight school's test scores are pulling the line up and to the right. Again this goes against what the school board member claimed.


### Task 3: Run the Regression Model

Let's build the linear regression model using all schools.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

There are three things to examine: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct.

The **p-value** tells you whether the relationship is statistically significant. The **Adjusted R-squared** tells you how much of the variation in math scores is explained by FARMS percentage — it's already a percentage, so a value of 0.025 means 2.5%, not 25%. The **coefficient** on FARMS Pct tells you the direction and size of the relationship: positive means higher FARMS predicts higher math scores; negative means higher FARMS predicts lower scores.

Report all three values and describe what they tell you. Does this model support the school board member's claim? Is this even a model worth using?

The p-value is 0.12, which is not statistically significant. We can't rule out that this relationship is just noise. The adjusted R-squared is 0.025, meaning FARMS percentage explains just 2.5% of the variation in math scores. And the coefficient is +0.12, so the model actually predicts higher FARMS is associated with slightly higher math scores. 

None of this supports the school board member's claim. The model isn't worth using: it's statistically insignificant, explains almost nothing, and points in the wrong direction.


### Task 4: What's Driving the Problem? Filter and Rerun.

In Task 1, you likely spotted schools with a FARMS percentage of 100. Those values are suspicious — in most districts, having *every* student on free/reduced lunch is unusual and may indicate a reporting artifact or data error. Let's see what happens when we exclude them.

```{r}
schools_filtered <- schools |> filter(`FARMS Pct` < 100)

model_filtered <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools_filtered)
summary(model_filtered)
```

Now compare the two models side by side:

```{r}
cat("Full dataset model:\n")
cat("  Coefficient:", coef(model)[2], "\n")
cat("  p-value:", summary(model)$coefficients[2,4], "\n")
cat("  Adjusted R-squared:", summary(model)$adj.r.squared, "\n\n")

cat("Filtered dataset model (FARMS < 100):\n")
cat("  Coefficient:", coef(model_filtered)[2], "\n")
cat("  p-value:", summary(model_filtered)$coefficients[2,4], "\n")
cat("  Adjusted R-squared:", summary(model_filtered)$adj.r.squared, "\n")
```

How did removing those schools change the model? Did the coefficient change direction? Did the p-value or R-squared improve meaningfully? What does this tell you about the influence of suspicious data values on a regression model?

When we get rid of the 100% FARMS schools, the coefficient jumps from +0.12 to +0.40 and the p-value dropped to 0.043, now technically significant. The adjusted R-squared rose from 2.5% to 6.5%. 

Since the coefficient keeps going in the positive direction we have further evidence that higher FARMS is associated with higher math scores, again contradicting the school board member. But we should note that the 8 schools are actually pull the coefficient toward zero which illustrates how just a small cluster of values, are distorting the regression. 


### Task 5: Even the Better Model Has Problems

Even after filtering, the Adjusted R-squared remains very low. This means FARMS percentage — even when the data is clean — explains very little of the variation in math scores across these schools.

Look at the filtered scatterplot:

```{r}
schools_filtered |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage (Excluding 100% FARMS Schools)")
```

With a low R-squared, the residuals from this model aren't very meaningful as labels of "over-performing" or "under-performing" — the model doesn't know enough to make that judgment reliably. But let's calculate them anyway to see the spread:

```{r}
schools_filtered <- schools_filtered |>
  mutate(
    predicted = predict(model_filtered),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools_filtered |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

What are the largest residuals? Given the model's weak R-squared, how much confidence would you have in labeling these schools as over- or under-performers? Why?

The largest residuals belong to Gulph Elementary (-36.7), Hereford Elementary (-32.5), and Whitemarsh Elementary (-30.8) on the underperforming side, and Woodland Elementary (+28.4) and Crooked Billet Elementary (+27.0) on the overperforming side. With an adjusted R-squared of just 6.5%, I wouldn't put much weight on any of these labels. A school flagged as an outlier here could easily be reflecting factors the model never measured, like teaching quality, curriculum, or neighborhood resources, rather than genuine over- or underperformance relative to poverty. Publishing these residuals as meaningful rankings would be misleading.


### Task 6: Write the Story Pitch

You've now run two models, examined the data quality and seen that even a cleaned-up model explains very little of the variation in math scores. Return to the school board member's claim: that schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch (3-5 sentences) that addresses the following: Does the data support or contradict the claim? What data quality issues did you find? What would you need - different data, a different model, additional reporting - before you could responsibly publish a story about this claim? Who would you call first?

The Free and Reduce Meals System is an essential program for hundreds of school kids across the country, including in Montgomery County, Pennsylvania schools. But a county school board member claimed that schools with less students enrolled in FARMS, a program that serves primarily low-income students, do better on standardized math tests. 

However, an analysis of district data doesn't support the school board members claim. School-level figures show no statistically significant relationship between FARMS rates and math scores, and even after cleaning out the 100% FARMS outliers, the adjusted R-squared is just 6.5%, meaning poverty rate explains almost none of the variation in outcomes. 

In fact, eight Montgommery County schools with a total student population using FARMS reported some of the highest Math test scores across the district.

My first call in this story would be to the school district, to get a better profile of the schools who tested high and also had their entire student population reportedly on FARMS. Of course, I'd want that FARMS figure fact checked by school administrators as well.
