---
title: "Polling Simulation: Sampling and the Central Limit Theorem"
---

In this activity, you'll use data from the 2024 Cooperative Election Study (CES) — a large-scale national survey conducted during the 2024 election — to simulate what happens when a news organization tries to estimate public opinion from a poll sample. The full dataset has nearly 55,000 respondents, which we'll treat as our population.

Your goal is to see firsthand how sample size affects the accuracy of estimates, and what the distribution of sample estimates actually looks like — the core idea behind the Central Limit Theorem.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from this URL and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

We'll focus on one question: what share of American adults identify as religiously unaffiliated (atheist, agnostic, or nothing in particular)? This has been a major story in American religion journalism for years.

First, we'll create a new column called `unaffiliated` that is `TRUE` if a respondent's `trad` value is `"Atheist"`, `"Agnostic"`, or `"Nothing in Particular"`, and `FALSE` otherwise. Then calculate the proportion of the full dataset that is unaffiliated. Save this value as `population_proportion`.

```{r}
ces <- ces |>
  mutate(unaffiliated = trad %in% c("Atheist", "Agnostic", "Nothing in Particular"))

population_proportion <- mean(ces$unaffiliated)
```

That population proportion is the "true" population value that any poll is trying to estimate. Write it down — you'll be comparing everything else to it.

### Task 4: Take a single small sample

Use `set.seed(42)` and then sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of unaffiliated respondents in your sample.

```{r}
set.seed(42)

sample_50 <- ces %>%  sample_n(50)

proportion_50 <- mean(sample_50$unaffiliated)
```

How close is your sample estimate to the original proportion? Is it surprising?

The original population proportion came back at about 36% unaffiliated. When I ran the sample at 50 we hit about 32%. Just 4 percentage below points the true value. Given the sample size I am not really surprised. A few outliers in the data would cause the percent to pull in any given way. 


### Task 5: Take a single larger sample

Now sample 500 rows. Set a different seed value, grab the sample and save it as `sample_500`. Calculate the proportion of unaffiliated respondents in this sample.

```{r}
set.seed(24)

sample_500 <- ces %>%  sample_n(500)

proportion_500 <- mean(sample_500$unaffiliated)

```

How did the estimate change compared to the sample of 50? What does this suggest about sample size?

I think this example shows us how an increased sample size can shift the quality of a projection. When we increase the sample to 500, we get back a 33.6%, which is closer to the true 36% than the n=50 estimate spit out previously. The gap shrank from 4 points to 2.4 points. Obviously it's not 36%, but the direction is exactly what we'd expect. With 10 times as many respondents, there's just less room for random quirks to push the estimate awy from the true proportion.

### Task 6: Simulate many polls

A single sample can get lucky or unlucky. To understand how polling actually works, we need to see the *distribution* of estimates across many samples. The code below draws 1,000 samples of size 50, calculates the proportion of unaffiliated respondents in each, and stores the results.

```{r}
set.seed(60)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$unaffiliated)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side — one for the 1,000 samples of size 50 and one for the 1,000 samples of size 500. Add a vertical line at `proportion` to each so you can see how well the samples are centering on the true value.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = population_proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share Religiously Unaffiliated",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**What shape do both histograms have?** Describe them in plain language. Does it surprise you that this shape emerges from a variable that is simply TRUE or FALSE for each respondent?

Both of the histograms are in the shape of a bell curve, which is surprising given a question with a binary answer like true or false. But the Central Limit Theorem tell us that the random "noise" of those individual 1s and 0s begin to cancel itself out as you take more samples. In other words, the average of each individual draw tends towards a normal distribution. 

**Compare the two histograms.** What is the most visible difference between the n=50 and n=500 distributions? What does that difference tell you about what happens when a poll uses a larger sample?

With a sample size of 500, the histogram is much narrower and taller than when the sample size is 50. The smaller sample size cause the data to fan out a bit wider in both direction, capturing how outliers in the data makes a bigger impact when there are fewer collections. A larger sample, doesn't just get us closer to the average, it also reduces how far off any single poll is going to be. A poll of 50 might land 10 or more percentage points from the truth, but a poll of 500 almost never does.

**Both distributions are centered near the red line.** What does it mean that 1,000 different random samples — each one drawn independently — tend to cluster around the same value? What would it mean for journalism if that *weren't* the case?

From what I can tell based on this example, the random sampling method seems to be relaiable for estimating the truth. Obviously no individual sample is perfect, but the process of randomized picks points consitently the right direction, in this case, across 1,000 trials. This is what makes polling during the political campaign season for news organizations. If the samples didn't center on the true value we could not meaningfully use this method since we would have no basis to trust any survey result irregardless of how large the sample. 

**A reporter tells you their story is based on a poll of 50 people.** Based on what you've seen in the simulation, what questions would you want to ask before trusting that number? You don't need to use any technical terms — just describe what you observed.

There are a few questions I'd like to raise, beginning with how the 50 people were chosen since it is such a small sample size and that can heavily affect the result in a way that a larger sample set wouldn't (again depending on how the people were selected for this poll). We already saw how wide the outliers can be from a randomized sample during our activity here, so I'd like to know more about the selection method for the poll (as in who took the poll). Without knowing whether this particular poll happened to be close or far off, the headline number alone doesn't tell the whole story.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
