---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: "Raphael Romero Ruiz"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

pct_unemployed is the strongest correlate of turnout at -0.74 — that's the biggest circle on the corrplot and it's not close. pct_white comes in second at 0.69, median_age third at 0.50, income fourth at 0.45. pct_bachelors_higher (0.30) and population (-0.39) are the weakest of the group. The scatterplots back this up: pct_unemployed has a steep, clean negative trend, pct_white has a clear positive one, and median_age tracks fairly linearly. Income trends positive but with more scatter, and population is mostly flat until the two big urban outliers drag the line down.

The collinearity concern is real but it's not between income and pct_unemployed — it's between income and pct_bachelors_higher, which correlate at 0.75. Those two are measuring too much of the same thing. I'd be skeptical of running both in the same model.


```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
* **Income**: Does higher income correlate with higher turnout?
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
* **Median Age**: Do counties with older populations have different turnout rates?
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
* **F-statistic**: Is the model as a whole statistically significant?


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

The adjusted R-squared is 0.80 against a raw of 0.85. That's not the overfitting mess I expected with six predictors and 24 counties. Three variables don't clear the bar: population (p=0.78), pct_bachelors_higher (p=0.84), and pct_unemployed (p=0.16). Income (p=0.005), pct_white (p=0.012), and median_age (p=0.04) are the ones that hold.

pct_unemployed at p=0.16 is the most interesting result in this output. It had the strongest bivariate correlation with turnout at -0.74 and it's insignificant in the full model. That's not a weak variable. That's a variable whose variance is already covered by pct_white, which it correlates with at -0.59. pct_bachelors_higher does the same thing: income absorbs it and it falls to p=0.84. I'd cut both and build from the three that held.

```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

Model 2 drops income to test whether education is the more useful of the two collinear variables or whether income was carrying the real weight. Model 3 goes down to median_age, pct_white, and pct_unemployed, the three that plotted cleanest against turnout. Model 4 swaps pct_white for pct_bachelors_higher to see if education predicts better than race with age and unemployment already in.

The full model wins at 0.80. Model 2 drops to 0.70, Model 3 to 0.62, Model 4 to 0.60. Dropping income alone cost 10 points of adjusted R-squared. That's not collinearity noise. That's income doing real predictive work that nothing else replaces. In Model 4, once income and pct_white are both out, pct_unemployed shoots to p=0.0003. That's the suppression flipping. The variables that were absorbing unemployment's variance in the full model are gone and it finally has room. I'd take the full model to my editor.

```{r}
# Model 2 - drop income to remove multicollinearity with education
model2 <- lm(turnout ~ population + pct_bachelors_higher + pct_unemployed + median_age + pct_white, data = md_turnout)
summary(model2)

# Model 3 - just the three strongest visual predictors
model3 <- lm(turnout ~ median_age + pct_white + pct_unemployed, data = md_turnout)
summary(model3)

# Model 4 - swap pct_white for pct_bachelors_higher
model4 <- lm(turnout ~ median_age + pct_bachelors_higher + pct_unemployed, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

The analysis points to a few counties worth calling first. Baltimore City at 0.58 and Prince George's County at 0.66 are both underperforming what their income and pct_white values would predict, which is a story about something structural that the demographic variables aren't capturing. That gap between predicted and actual turnout is where I'd start reporting.

The age finding gives me a specific frame. Maryland's oldest counties, Talbot and Kent, are among the highest-turnout jurisdictions in the state. I'd want to know whether that's a composition effect — more older residents — or whether older residents actually show up at higher rates. Those are two very different stories with two very different reporting angles. My first calls would be to election administrators in both the high-performing and underperforming counties.
