---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_songs <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
mean_danceability <- mean(spotify_songs$danceability)
sd_danceability <- sd(spotify_songs$danceability)

mean_danceability
sd_danceability
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
hist(spotify_songs$danceability)
```

### Task 5: Add a line to the histogram for the mean danceability score

Write code to add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
hist(spotify_songs$danceability) 
abline(v = mean_danceability, col = "red", lwd = 2)
```

### Task 6: Describe our results

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and distribution tell us about the songs in the dataset?

The 2020 Spotify dataset shows us that danceability scores are notably left-skewed, which means there were more songs with higher danceability with just a few that scored lower. With an average score of 0.67, it’s clear that most tracks are more on the upbeat side. The standard deviation of 0.16 indicates that there’s a moderate amount of consistency, with many songs clustering in the higher range of danceability. Overall, this suggests that energetic and rhythmic music was a big favorite among listeners and dominated the charts in 2020. I guess the pandemic had us all dancing in our living rooms!

### Task 7: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

