---
title: "HW7b - Probability and the Normal Distribution"
---

In this assignment, you'll build on your z-score skills to answer probability questions using the normal distribution. You'll return to the White House salaries data and use it to answer the kinds of "what are the chances?" questions that come up in journalism all the time. This sets the stage for sampling, which we'll cover next.

### Task 1: Load the tidyverse and the data (1 point)

Load the tidyverse and the White House salaries data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv into a dataframe called `wh_salaries`

```{r}
library(tidyverse)

wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```

### Task 2: Filter and summarize (2 points)

You may remember from HW3 that some White House staffers have a salary of $0. Let's filter those out and focus on paid employees, saving the result to a new dataframe called `wh_paid`. You need to calculate the mean, median and standard deviation of the salary column using `wh_paid`:

```{r}
wh_paid <- wh_salaries |>
  filter(SALARY > 0)

wh_paid |> summarize(
  mean_salary = mean(SALARY),
  median_salary = median(SALARY),
  sd_salary = sd(SALARY)
)
```

Now create a histogram of the paid salaries with a vertical line for the mean and two more for ±1 standard deviation, similar to what you did in HW6 with Spotify data.

```{r}
library(ggplot2)

mean_sal <- mean(wh_paid$SALARY)
sd_sal <- sd(wh_paid$SALARY)

ggplot(wh_paid, aes(x = SALARY)) +
  geom_histogram(bins = 30, fill = "lightblue", alpha = 0.7) +
  geom_vline(xintercept = mean_sal, color = "red", linetype = "dashed") +
  geom_vline(xintercept = mean_sal + sd_sal, color = "orange", linetype = "dashed") +
  geom_vline(xintercept = mean_sal - sd_sal, color = "orange", linetype = "dashed")
```

**Describe the shape of the distribution. Is it roughly normal? Skewed? Where do most salaries cluster?**


The histogram shows that this distribution is definitely not normal, but instead, bimodal, meaning it has two distinct peaks. Most salaries cluster in two very different spots: a lower-tier group between 60k and 75k and on the middle income earners at around 120k to 130k. These likely represent the split between administrative support and more senior policy or legal staff.

Notably there is also a third, smaller spike leading up to the 200k range. This has to be where the senior-most advisers and political appointees sit, likely hitting a federal salary cap. The mean value sits right in the middle of the two big clusters at 111,322. The typical person working here earns 101,500 which is about 10k lower than the average. This confirms what we see in the shape in the histogram. The distribution is not symmetrical in this data.



### Task 3: The Empirical Rule (3 points)

The empirical rule (also called the 68-95-99.7 rule) says that for a normal distribution:
- About 68% of data falls within 1 standard deviation of the mean
- About 95% falls within 2 standard deviations
- About 99.7% falls within 3 standard deviations

Let's test this. We'll calculate the salary ranges for 1, 2, and 3 standard deviations from the mean, then count how many employees actually fall in each range.

```{r}
mean_sal <- mean(wh_paid$SALARY)
sd_sal <- sd(wh_paid$SALARY)
total_n <- nrow(wh_paid)

empirical <- tibble(
  range = c("Within 1 SD", "Within 2 SD", "Within 3 SD"),
  lower = c(mean_sal - sd_sal, mean_sal - 2 * sd_sal, mean_sal - 3 * sd_sal),
  upper = c(mean_sal + sd_sal, mean_sal + 2 * sd_sal, mean_sal + 3 * sd_sal),
  expected_pct = c(68, 95, 99.7),
  actual_count = c(
    sum(wh_paid$SALARY >= (mean_sal - sd_sal) & wh_paid$SALARY <= (mean_sal + sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 2 * sd_sal) & wh_paid$SALARY <= (mean_sal + 2 * sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 3 * sd_sal) & wh_paid$SALARY <= (mean_sal + 3 * sd_sal))
  ),
  actual_pct = round((actual_count / total_n) * 100, 1)
)

empirical
```

**How closely does the data follow the empirical rule? What does any deviation tell you about the distribution?**

The data roughly follows the empirical rule. Within 1 standard deviation more than half of the employees, 65.9%, fall in range versus the expected 68%. The bigger divergence comes at 2 SDs, where only 91.2% of salaries fall within range instead of 95%. And all of the employees, literally 100%, fall within 3 SDs.

The histogram showed us a bimodal distribution and the deviations support this. The tail of the data doesn't taper off and instead there's that small spike we mentioned before at 2 SDs.

### Task 4: From z-scores to probability with pnorm() (4 points)

Remember the car thefts z-score activity from class? Well, I want to return to that briefly, introduce a new tool called pnorm() and then apply it here. So open that CarTheftZScores notebook and, at the bottom, copy and paste in the contents here: https://gist.githubusercontent.com/dwillis/25102889f576b1707576b85e0be0c353/raw/1ca2be9636497cb5293ff7b1f4e6ed8e27a9e049/car_theft_pnorm.rmd

Re-run all of the blocks in CarTheftZScores notebook, including the code you just pasted in. What's happening there is you are using a function called `pnorm()` to find the probability of a value being less than or equal to a certain z-score. The basic formula to find that probability is:

```
1 - pnorm(z_score)
```
What I saw after rerunning the CarTheftZScores
Month: 2023-02-01
Total thefts: 685

Z-score: 2.81
pnorm() estimate - probability of a month this high or higher: 0.0025 (0.25%)
Actual data - proportion of months this high or higher: 0.0103 (1.03%)

Where `z_score` is one row's z-score. Now apply the same approach to salaries.

Your editor asks: **"What's the probability that a randomly selected White House employee makes more than $150,000?"**

To answer this:
1. Calculate the z-score for $150,000
2. Use `pnorm()` to find the probability. `pnorm()` gives you the area under the normal curve to the LEFT of a z-score — that is, the probability of being BELOW that value.
3. Subtract from 1 to get the probability of being ABOVE $150,000
4. Check your answer against the actual data

```{r}

salary <- 150000
z_score <- (salary - mean_sal) / sd_sal

prob_less <- pnorm(z_score)
prob_more <- 1 - prob_less
prob_more

# Check against actual data
actual_prob <- sum(wh_paid$SALARY > 150000) / total_n
actual_prob
```

**Write a sentence describing this probability in plain English. How does the normal approximation (prob_less) compare to the actual percentage?**

Based on the actual data, about 18.2% of paid White House employees,or one in six, earn more than 150k a year. The normal distribution model estimated this at 16.9%, which underestimates the true figure by about 1.3 percentage points.

That gap exists because, as we saw in the histogram, the bimodal salary structure has that cluster of senior staffers near the 200k range. The normal curve gives us a very similar figure albeit a more conservative estimate than what is actually happening. Obviously we would want to report the actual figure. If we are going based off of the normal model, it may cause us to narrow our scope of investigation on who and what kind of jobs merit such an uncommonly high salary and likely the people we should be holding to a higher scrutiny.

### Task 5: Probability between two values (3 points)

Now your editor asks: **"What percentage of White House employees make between $80,000 and $120,000?"**

You'll need to calculate z-scores for those two figures, use `pnorm()` for each, and subtract to find the probability between them. Update the REPLACE_ME figures with the appropriate numbers.

```{r}

z_lower <- (80000 - mean_sal) / sd_sal
z_upper <- (120000 - mean_sal) / sd_sal

prob_between <- pnorm(z_upper) - pnorm(z_lower)
prob_between

# Verify with actual data
actual_between <- sum(wh_paid$SALARY >= 80000 & wh_paid$SALARY <= 120000) / total_n
actual_between

```

**Interpret this result. What does this "middle range" tell us about the White House salary structure?**

About 32.6% of White House employees, or roughly 1 in 3, earn between 80k and 120k. The normal distribution overestimates this at 36.6%, a gap of 4 percentage points. The normal curve expects a larger number of employees to cluster around the average, at 111k, but that's not what we see in the histogram. Most of the people working at the White House are getting paid either in the lower-earner or higher-earner brackets, a pay gap that could signal to us reporters about how leadership structures exist within the staffers. If so few employees fall between 80,000 and 120,000, that gap itself becomes the story since it may be that there are lots of junior staffers keeping the operation together in the White House and power being concreted amoung only a few decision makers. 

### Task 6: Expected vs. observed (3 points)

Let's compare what the normal distribution *predicts* to what we actually *see*. This idea — comparing expected to observed — is one you'll use again later in the course.

Create these salary bins and calculate the expected count (from the normal distribution) and the observed count (from the actual data) for each:

```{r}
bins <- tibble(
  category = c("Below $60k", "$60k-$90k", "$90k-$120k", "$120k-$150k", "Above $150k"),
  lower = c(-Inf, 60000, 90000, 120000, 150000),
  upper = c(60000, 90000, 120000, 150000, Inf)
) |>
  mutate(
    expected_pct = pnorm(upper, mean = mean_sal, sd = sd_sal) - pnorm(lower, mean = mean_sal, sd = sd_sal),
    expected_count = round(expected_pct * total_n, 1),
    observed_count = c(
      sum(wh_paid$SALARY < 60000),
      sum(wh_paid$SALARY >= 60000 & wh_paid$SALARY < 90000),
      sum(wh_paid$SALARY >= 90000 & wh_paid$SALARY < 120000),
      sum(wh_paid$SALARY >= 120000 & wh_paid$SALARY < 150000),
      sum(wh_paid$SALARY >= 150000)
    ),
    difference = observed_count - expected_count
  )

bins |> select(category, expected_count, observed_count, difference)
```

**Where does reality differ most from what the normal distribution predicts? What might explain those differences?**

The normal model predicted about 40 employees earning below 60k, but only 2 actually do. This suggests that there could be a salary floor at the White House which we would not see if we applied a normal curve since it would smooth that low count. It also predicts a lower count in that lower-earner range we have been discussing. It again tries to paint a picture of a more equally spread out earning range. But the reality is that there is a bigger gap from here and the next two ranges. The 90k–150k range has about 49 people combined less than what the model expected. This again reinforces my take that the middle managers at the White House are absent. At more than 150k, we see that there are 14 more employees than expected, which aligns with that cluster of senior advisers hitting a pay cap, likely.

### Task 7: Story pitch (4 points)

Based on your probability analysis, write a headline and lede for a story about White House salaries. Your lede should incorporate at least one probability or percentage you calculated.

Then write 2-3 sentences describing how you would use probability thinking to advance this story — what could you compare these probabilities to, and what additional analysis would help?

Headline: 1 in 5 White House Staffers Out-Earn the Average American by More Than Double

Lede: While the average American full-time worker brought home $63,360 in 2025, roughly 1 in 5 White House employees cleared $150,000 — more than double the national median — according to an analysis of White House payroll disclosures."

Nut graf: The payroll data shows that those high earners aren't evenly scattered — they cluster near 195,000, the federal statutory salary cap, pointing to a concentrated group of senior political appointees and advisers driving the top-tier numbers. Meanwhile, the largest single group of White House staff earns between 60,000 and 90,000, far more than a normal salary curve would predict. The result is a payroll with two distinct poles and a hollowed-out middle — raising questions about who gets a seat at the table and at what price.

Advancing the Story
To advance this story, I'd compare these salary figures to what prior administrations paid — to see if this concentration of high earners is actually new. I'd also compare those salaries to what federal employees doing similar jobs earn at other agencies, to see if the White House is simply paying a premium for proximity to power. Finally, a closer look at job titles in the payroll data would show exactly who's landing those six-figure salaries and whose position vanished from the middle.\

I'd also like to drive the point home further about the growing distance among the average American worker and the politco class at the helm of the White House by getting salary estimates for jobs that people who voted for Trump hold. 

### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
