---
title: "HW8b: Polling Simulation"
---

In class, you simulated polling estimates for the share of Americans who are religiously unaffiliated — a group that makes up about 36% of the population. In this assignment, you'll run the same simulation for a different group: White Evangelicals, who make up roughly 10% of the population.

That difference matters more than it might seem. When a proportion is close to 50%, samples tend to be pretty well-behaved. When it's much smaller — or much larger — samples can behave differently, especially at small sizes. Your job is to run the simulation, compare what you find to the in-class results, and think carefully about what it means for reporting on religion and politics.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a respondent's `trad` value is `"White Evangelical"` and `FALSE` otherwise. Then calculate the proportion of the full dataset that is White Evangelical and save it as `population_proportion`.

```{r}
ces <- ces |>
  mutate(evangelical = trad == "White Evangelical")

population_proportion <- mean(ces$evangelical)
population_proportion
```

Write down this value. How does it compare to the unaffiliated proportion (about 0.36) from the in-class activity?

We have a smaller percentage of White Evangelicals, at approximately 0.10 or 10% of the respondents, compared to the unaffiliated proportion from our in class activity. Our population here is about one-third the size which matters a lot in how we poll since we know that groups with fewer members in any given sample make estimates much noisier and less reliable at the same sample sizes. 


### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of White Evangelical respondents in your sample.

```{r}
set.seed(42)
sample_50 <- ces |> sample_n(50)
mean(sample_50$evangelical)
```

How close is your sample estimate to the population proportion? How does the accuracy compare to what you saw with the unaffiliated group in class?

With n at 50 we'd expect that about 5 White Evangelical respondents would show up in our sample since we know that our true proportion is about 10%. Here we see we actually got a tad more than the true proportion, at 0.16 or 16%, which would translate to 8 respondents. This is not all that surprising since the sample size is small enough that we could see swings in our estimate that would cause our percentage to range a few percentage points. Compared to the unaffiliated group in our in class exercise the improvement is not as night and day since the population we are focusing is so rare in to begin with.

### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`. Calculate the proportion of White Evangelical respondents in this sample.

```{r}
set.seed(1966)
sample_500 <- ces |> sample_n(500)
mean(sample_500$evangelical)
```

Did the larger sample improve the estimate? How does the improvement compare to what you saw in class?

With a sample size at 500 our accuracy is much improved. This sample produces about 50 White Evangelical respondents, which is much closer to the true proportion we found a few chunks above. The improvement mirrors what we saw in class for the unaffiliated group, again reinforcing that sample size is incredibly important. Compared to our in class assignment the jump here is more consequential since the reliability at the sample size is greatly improved. 

### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but using your `evangelical` column. Run it as-is.

```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class. Add a vertical line at `population_proportion`. Make sure to update the title and subtitle to reflect that this simulation is about White Evangelicals.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |>
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = population_proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Scaling up the sample size stabilizes White Evangelical estimates",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the unaffiliated histogram from class? Why do you think that is, given that White Evangelicals make up only about 10% of the population?

With n at 50, the distribution is noticeably right-skewed and lacks the symmetry we have seen in more balanced datasets. The histogram piles up near the zero mark with a long tail stretching toward the right. This skew occurs because the true population proportion, which is about 10%, is seated right against the our zero floor. In a small sample of 50, we would again expect about five evangelical respondents. But if a few of those are missing, the result crashes into the floor. There is also lots of room for a lucky draw to pull the proportion toward 20% or higher by chance, hence the tail. While the Central Limit Theorem eventually smooths data into a bell curve, it struggles at this scale when the underlying reality is this close to zero. The sniff test here tells us that at 50, our estimates is more likely to be impacted by random variation than the 10% true proportion.

**Compare the n=50 and n=500 histograms for this group.** How much does the distribution change with a larger sample? Is the improvement bigger or smaller than what you saw in class with the unaffiliated group?

From 50 to 500 we see a big improvement visible in our histogram with a narrower, symmetrical chart and a center that is right about the true proportion. Like I said the change makes an bigger difference to this subgroup compared to the unaffiliated group from our in class assignment. There was more stability bumping the sample size to 50 with the unaffiliated group in a way that we just don't see in our homework's group. The change here basically fixed our shape problem, making it much more reliable due to the reducing margin of error. Once again driving home the lesson that more people equals better data in a polling context.


**A news organization wants to report on White Evangelical voters using a poll of 200 people.** Based on your simulations, roughly how many White Evangelical respondents would that poll be likely to include? What does that suggest about the reliability of estimates for smaller subgroups?

In a 200-person poll I would expect it to include only about 20 White Evangelical respondents. This seems like too small of a population to do any real serious reporting, unless of course we are localizing a story and are looking at say a town's evangelical population. Though still, at this scale, each individual respondent represents 5 percentage points of that subgroup's total meaning that if just three or four people answer differently the estimate would swing. I would not rely on smaller subgroup estimates like this due to the noisy-ness, at least at the 200 size. We need to increase it.

**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** 

# Margin of error for White Evangelical subgroup in a 400-person poll

```{r}
n_total <- 400
pop_proportion_evangelical <- 0.10  # ~10% of population

n_evangelical <- n_total * pop_proportion_evangelical  # expected subgroup size
sup_percent <- 0.78  # the support figure

moe <- 1.96 * sqrt((p * (1 - p)) / n_evangelical)
```


**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** The poll sampled 400 people. Based on everything you've seen in this assignment and in class, what would you want to know before reporting that number?

Before this hits the wire, it needs a heavy sniff test. First, I want to know the raw count of White Evangelicals in that sample, this is probably the most consequential variable again due to the rarity of the population we are looking to understand. At 10%, that’s only about 40 people, meaning the margin of error for that subgroup is going to be pretty big especially in the context of an election. In this case I calculated it above at just shy of 13 percentage points in either direction. Yikes.  Second, would be a quality test on the the question prompting a response. What do they mean by support? Was the question framed in a way that could be problematic? Are we leading people on to a "supportive" response? I'd also want to know how the group was defined. Finally, I'd check if the sample was actually representative or if it was a lucky draw of high-support outliers. Without those details, that 78% is just a noisy number that tells us very little about the actual electorate. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
