---
title: "HW9: Car Theft Confidence Intervals & Margin of Error"
name: Raphael Romero Ruiz
---

Your editor wants to know what a "typical" month of car thefts looks like in Prince George's County. Not just the average — she wants to know how confident you can be in that number. Using monthly figures, confidence intervals and margin of error, you'll come up with an answer and decide whether the data supports a clear story. As with the previous homework, provide answers in this notebook when prompted and you will have to fill in values where you see REPLACE_ME. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### Task 2: Load the Prince George's county crime data

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)

```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. What's your initial reaction to the change in car thefts? Is there a trend? If your editor asked you "how many car thefts happen in a typical month?", what would you say based on this chart alone?

There is a clear and strong downward trend in car thefts in Prince George's County. The data starts very high — over 600 thefts in July 2023 — and declines fairly steadily to around 200–280 by late 2025 and early 2026. There are some bumps and plateaus along the way (notably mid-2024), but the overall direction is unmistakably downward. Based on the chart alone, I'd tell my editor that a "typical" month is hard to pin down because the number is moving so much — somewhere between 350 and 425 if you average across the whole period, but recent months are running closer to 200–300.

```{r}

car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2026)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```

### Task 4: Calculate the mean, standard error and margin of error

We've got a small number of months, so we need to quantify the uncertainty around our average. We'll calculate the standard error (which measures how much the sample mean could vary), and the margin of error (which gives us the plus/minus range). Replace the REPLACE_ME values based on what we did in class.

```{r}

theft_summary <- car_thefts |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts
  )

# Display the results
theft_summary

```

**Answer this question:** Explain in your own words what the margin of error tells us here. If you were writing a story, how would you use this number? For example, how would you complete this sentence: "Prince George's County averages about ___ car thefts per month, plus or minus ___."

The margin of error tells us how much the sample mean could plausibly vary if we were drawing this estimate from a larger population of months. In practical terms, it gives readers a sense of how precise — or imprecise — our average really is. Rather than stating a single number as definitive fact, the margin of error signals a range of credibility. In a story, I would use it to be honest about uncertainty: "Prince George's County averages about 376 car thefts per month, plus or minus 34." That means the true typical month likely falls somewhere between 342 and 410 — a meaningful range that matters when evaluating policy claims or crime trends.

### Task 5: Build the confidence interval

Now let's turn the margin of error into a full confidence interval — the range where we're 95% confident the true average falls. Replace the REPLACE_ME values.

```{r}

theft_summary <- theft_summary |>
  mutate(
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

theft_summary |>
  select(mean_thefts, margin_error, ci_lower, ci_upper)

```

**Answer this question:** If a county official claimed "car thefts are down to about 300 per month," does your confidence interval support that claim? What about a claim of 350 per month? Explain how the confidence interval helps you evaluate each claim.

The confidence interval (approximately 341 to 410 per month) does not support the claim of 300 per month — that number falls below the lower bound of our interval, meaning we can be reasonably confident the true average is higher than 300. The claim of 350 per month, however, does fall within our confidence interval, so the data is consistent with that figure. The confidence interval is useful here because it gives me a principled range to evaluate official claims against: if a number falls outside the interval, I have statistical grounds to push back; if it falls inside, I can't definitively rule it out.

### Task 6: Visualize the data with the confidence interval

Let's make a chart showing the data with the mean and the 95% confidence interval. Does this change anything about your answer from Task 3?

Yes, the CI visualization reinforces and sharpens my Task 3 answer. The mean (~376) sits squarely in the middle of the early-to-mid portion of the data, but the chart makes clear that nearly every month from mid-2025 onward falls below the lower bound of the confidence interval. That's a strong visual signal that the current reality is well outside what the historical average predicts. The CI is still centered on the overall mean, so it's technically accurate over the full period — but relying on that average to characterize "today" would be misleading. The trend matters as much as the mean.

```{r}
# Make a chart with the mean and CI
car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper),
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month),
           y = theft_summary$mean_thefts + 10,
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)),
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 7: The margin of error and sample size

In class we learned that MoE = 1/√n × 100 as a quick approximation for polls. That same principle applies here: more data means a smaller margin of error. Let's see what would happen if we had 40 months of data instead of the number we have, assuming the same mean and standard deviation. Replace the REPLACE_ME value with 40.

```{r}

hypothetical_summary <- tibble(
  mean_thefts = theft_summary$mean_thefts,
  sd_thefts = theft_summary$sd_thefts,
  months = 40,
  se_thefts = sd_thefts / sqrt(months),
  margin_error = qt(0.975, df = months - 1) * se_thefts,
  ci_lower = mean_thefts - margin_error,
  ci_upper = mean_thefts + margin_error
)

# Compare the two
comparison <- tibble(
  scenario = c("Actual data", "40 months"),
  months = c(theft_summary$months, hypothetical_summary$months),
  margin_error = c(theft_summary$margin_error, hypothetical_summary$margin_error),
  ci_lower = c(theft_summary$ci_lower, hypothetical_summary$ci_lower),
  ci_upper = c(theft_summary$ci_upper, hypothetical_summary$ci_upper)
)

comparison

```

**Answer this question:** How did the margin of error and confidence interval change? What does this tell you about the tradeoff between having more data and being able to make precise claims? If you were covering crime trends, how many months of data would you want before writing a definitive story?

With 40 hypothetical months, the margin of error shrinks from about 34 to about 30, and the confidence interval narrows accordingly. The change isn't dramatic, but it illustrates a real principle: every additional data point reduces uncertainty, though with diminishing returns (since MoE shrinks by the square root of n, not n itself). The tradeoff is that more data takes time to collect — meaning journalists covering an emerging trend may have to publish with wider uncertainty and be more careful about hedging. For crime trends, I'd want at least 24–36 months (2–3 years) before writing a definitive story. Anything less is too susceptible to seasonal spikes or single unusual events distorting the picture.

### Task 8: Reporting with uncertainty

Based on everything you've calculated, answer the following:

1. Write a headline and a one-paragraph lede about car thefts in Prince George's County. Your text should properly reflect the uncertainty in the data — don't overstate or understate what the numbers show.

2. Your editor says "Just tell me if car thefts are going up or down." Using only the confidence interval (no significance testing yet), what's the most accurate answer you can give? What would you need to do next to answer that question more definitively?

**Headline:** Car Thefts in Prince George's County Have Fallen Sharply — But Uncertainty Remains About How Low They'll Go

**Lede:** Car thefts in Prince George's County have dropped dramatically since mid-2023, falling from more than 600 incidents a month to roughly 200–280 in recent months, according to county crime data. While the county averaged about 376 thefts per month over the past two and a half years — a figure that could reasonably range from 342 to 410 given the sample size — the most recent months suggest the actual current rate may be well below that historical average, though analysts caution that a longer data record would be needed to confirm the trend is permanent.

**Editor's question — "Are car thefts going up or down?":** Based purely on the confidence interval, the most accurate answer is: they appear to be going down, but the CI alone can't prove it definitively. The overall CI (341–410) is anchored to the full 32-month average, and it captures the middle range of the data — not the most recent trajectory. To answer the question more definitively, the next step would be a significance test (such as a linear regression or a test comparing early-period vs. late-period means) to determine whether the observed decline is statistically significant and not just noise.


When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
